### 2021-12-07 manuscript.tex
Add a preface [ci skip]
This build is based on
https://github.com/lubianat/quali_phd/commit/e6cf449283aa647015cb56637d7c2de7d9f53a02.

This commit was created by the following CI build and job:
https://github.com/lubianat/quali_phd/commit/e6cf449283aa647015cb56637d7c2de7d9f53a02/checks
https://github.com/lubianat/quali_phd/actions/runs/1551216151

This manuscript
(\href{https://lubianat.github.io/quali_phd/v/856180b09ff9ec0998a2a1d9e24bc3364473fbc9/}{permalink})
was automatically generated
from \href{https://github.com/lubianat/quali_phd/tree/856180b09ff9ec0998a2a1d9e24bc3364473fbc9}{lubianat/quali\_phd@856180b}
on December 6, 2021.

@@

The Human Cell Atlas (HCA) is an international effort aiming at characterizing every cell type of the human body.
By the virtue of tecniques such as single-cell RNA sequencing, mass cytometry, and multiplexed in situ hybridization, HCA members are producing cell-level data from virtually all human tissues.
This wealth of data can have a significant impact on biomedical research, but only if its content is genuinely interoperable.
While ontologies and semantic technologies have emerged as key players in the data interoperability ecosystem, there are still gaps to cover between the technical possibilities and the practical applications in biomedical research.
Wikidata is a knowledge graph database emerging as a FAIR (Findable, Accessible, Interoperable and Reusable) repository for biological knowledge.
The formatting and deployment of information from the Human Cell Atlas to Wikidata can increase information availability and impact, by inserting the findings in a network containing multiple associations of concepts of all areas of knowledge (within and outside science).
Conceptually defining cell types in a general and applicable concept, formalized into a database-compatible format, is a massive theoretical challenge.
This PhD project aims at studying our current understanding of cell types for development a comprehensive ontological model in Wikidata for cell types.
We will review the single-cell literature, refining and formalizing concepts for cell type delimitation. Furthermore, we will use Natural Language Processing and Machine Learning tools to automate knowledge extraction from scientific articles in the scope of the Human Cell Atlas.
In an advanced step, we will apply concepts of network theory to develop tools for user-friendly querying of the database, making the knowledge ready for the academic community.

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

\hypertarget{introduction-1}{%
\section{Introduction}\label{introduction-1}}

@@

\hypertarget{literature-based-discovery}{%
\subsection{Literature Based Discovery}\label{literature-based-discovery}}

The amount of scholar information vastly outnumbers what single researchers can fathom.
Nevertheless, the gap between single individuals and the collectively body of knowledge has been widening in
an accelerated fashion.
The explosion in the number of published articles is leading to a ``tsunami of knowlegde'',
flooding the scientific literature with rich information.
That trend became spacially clear during the COVID-19 pandemic, when the huge amount of research published made keeping up with the literature pratically impossible. {[}\protect\hyperlink{ref-oVsnSEwQ}{61}{]}
At the same time, articles themselves are becoming denser, as high-throughput (and high-information) technologies
like single-cell RNA-sequencing get cheaper and widely used.
In practice, thus, too much of the knowledge generated remains unseen by any individual researcher, limiting the reach of science in general.

The technological advances, however, are no yet met by equivalent knowledge-handling systems.
Mainstream scientific publication is, nowadays, barely readable by machines.
Articles are written for human consumption, using ambiguous natural language and relying on implicit conventions.
Tables and data rarely make use of technical standars, such as employing URIs (Uniform Resource Identifiers or encoding information in RDF formats.
In fact, those standards and their acronyms are foreign for most life scientists (personal observations), despite
being the \emph{de jure} gold standard for data quality. {[}\protect\hyperlink{ref-UI1dtNSw}{62}/{]}
Interconnecting biomedical knowledge is an open challenge of our century, and there is a large way to go
before society can fully benefit from the sum of all knowledge we generate.

The scientific community has pursue solutions for this tsunami of information from many different angles.
Narrative reviews, systematic reviews and textbooks compile and synthetize information, providing a layer of processing.
Biocuration efforts go a step further and transform unstructured information into structured information
in knowledgebases, such as UniProt {[}\protect\hyperlink{ref-1F9v68m9B}{63}{]} and PDB.
Text-mining apply a range of Natural Language Processing tools to try and extract biological relations,
or provide guidance for biocurators.
Elaborate knowlegde networks, like the STRING database {[}\protect\hyperlink{ref-POzqtz7B}{64}{]} and Wikidata{[}\protect\hyperlink{ref-3GqlN9Dk}{65}{]},
combine information from different sources.

The synthesis effort of literature mining is not only an exhibition of the scientific claims in the literature.
Interconnected knowledge provides a way to discover new, implicit knowledge, by applying logical reasoning to a dataset.
A field denominated Literature Based Discovery {[}\protect\hyperlink{ref-jXryyPLB}{66}{]} dedicates itself to this challenge: make actual discoveries (or at least very strong hypothesis) using as material plainly the existing literature. {[}\protect\hyperlink{ref-Fuf8oc85}{67}{]}
The textbook example of Literature Based Discovery is described by Don Swanson's ABC model: If A is related do B, and B is related to C, then A and C are indirectly related {[}\protect\hyperlink{ref-LNtevqb9}{68}{]}.
In a seminal paper, Swanson showed an hypothesis about using fish oil (A) to treat Raynauld's disease (C), demonstrating that even though the specialized fish-oil (A) literature had shown its association (AB) with a set of blood parameters (B), and the specialized Raynauld's disease literature had show its association (BC) with the same set of parameters (B), the AC link was never made in the literature, despite its seeming obviousness {[}\protect\hyperlink{ref-LNtevqb9}{68}{]}

Modern advancements of literature-based discovery rely on Natural Language Processing, Machine Learning and Knowledge graphs to make inferences on literature knowledge.
Word embeddings, for example, are leading inference of properties of compounds based on their shared neighbourhood of words (the words before and after their mentiongs) with known compounds, thus making use of latent knowledge in the body of knowledge. {[}\protect\hyperlink{ref-Lbqx9XBK}{69}{]}
Other, more explicit approaches, rely on extracted relations embedded in knowledge graphs. As an example, the discovery of new RNA-binding proteins related to Amyotrophic Lateral Sclerosis by analysis of the Watson Drug Discovery gene-disease network. {[}\protect\hyperlink{ref-CovO24JR}{70}{]}

Knowledge graphs have a set of characteristics that make then useful for Literature Based Discovery: they represent multiple relations, allow inferences on top of those relations, and provide human understandability at every step, allowing for a dialog between expert humans and computing systems.
The field of biomedical ontologies explores that direction in depth, and the community is building many solutions, widely applicable for the biomedical sciences.

For the Human Cell Atlas Project (as presented in the chapter ) to maximize its benefit for society, it is arguably important that its knowledge products are inserted into the main route of automated knowledge discovery .
That implies a task of building knowledge graphs able to deal with it at all layers, including the generated data and metadata, its range of different protocols, and the purified knowledge projects that are enshrined in publications.
Thus, the chapter will present challenges and paths for applying literature based discovery on a large scale and with sufficient flexibility to deal with the Human Cell Atlas.

@@

The classification of biological concepts is at the core of biology. At least since the Aristotelian endeavours to group classes of animals, a good part of the scientific work is to capture concepts into knowledge systems {[}\protect\hyperlink{ref-DLTMbf54}{71}{]}.
Linnaeus' binomial system for naming species and Mendeleiev's periodic table are likely the two most famous classification systems, but are part of a much larger ecosystem of structuring scientific knowledge.

On the 20th century, the development of the analytical philosophy of Russel and Wittgenstein and their search for formalizations {[}\protect\hyperlink{ref-cZEmYsqx}{72}{]} gradually layed the foundations for the the logic of scientific descriptions. Karl Popper and his ``The Logic of Scientific Discovery''{[}\protect\hyperlink{ref-to2onCDy}{73}{]} was heavily influenced by analytical philosophy, and the field is at the foundation of the ``falseability'' system of Popper. Less known among life scientists, Tarski's inquiries on what can be considered to be ``true'' {[}\protect\hyperlink{ref-RghcuiSS}{74}{]} were also

@@

An ontology, as used here, is a formal computational representation of reality, which tries to represent each concept (and their relations) as precisely as possible. {[}\protect\hyperlink{ref-DLTMbf54}{71}{]}\\
Constructing an ontology is a process of selecting and defining terms of interest, selecting and defining relationships of interest and making statements about reality using terms and relationships.
The Gene Ontology is probably the most well known biomedical ontology; it describes (among other things) different classes of biological process, related to each\_other by ``is\_a'' and ``part\_of relations. {[}\protect\hyperlink{ref-18wkLcGxT}{75}{]} {[}\protect\hyperlink{ref-2KmDs8gy}{76}{]}.

The Gene Ontology is part of a much larger effort to formalize concepts across biology: the Open Biomedical and Biological Ontologies (OBO) Foundry. {[}\protect\hyperlink{ref-j3tJyG7J}{77}{]}
Created in 2007, the OBO Foundry is a hub of biomedical ontologies that sets guidelines for the design and construction of high-quality ontologies.
The initial OBO Foundry united several independent ontologies, like the Cell Ontology (CL), the Disease Ontology (DO) and the Protein Ontology (PRO) under a common framework towards interoperability.
At the same time, the creation of the Relation Ontology (RO) provided a go-to point for relations in biology that could them be reused by different ontologies.

@@

The Google Knowledge Graph introduced the Semantic Web \emph{de facto} in the daily life of users of Google. {[}\protect\hyperlink{ref-16mCkZC1r}{78}/{]}. Its underlying structure is similar to the triples in an ontology, but it is less concerned with being logically coherent, and does have strict semantics of a representation.
In that way, Google Knowledge Graphs can feed on a variety of sources and not crash if there is some data modelling that, rigorously, could be inconsistent.
Even though there is not a rigorous boundary between ontologies and knowledge graphs, one reasonable interpretation is that a knowledge graph may not be perfectly coherent, as long as it still can provide enough knowledge and reasoning for the approach of interest.
While the lack of formal semantics limits reasoning and inference, the knowledge graphs are arguably easier to use, edit and and understand, and so provide an user friendly alternative for computable information with a lower entry barrier.

While the Google Knowledge Graph is widely used as a source of knowledge, it does not allow independent users to contribute with information.
On the other hand, Wikidata, the collaborative knowledge graph of the Wikimedia foundation, allows users to contribute with classes and statements, in the same spirit of Wikipedia and share its ``epistemic virtues, like power, speed and availability. {[}\protect\hyperlink{ref-qMozvNth}{79}{]}
Its power is derived of its large community of contributors, closedly linked to the hugely successful Wikipedia.
With a community of more than 20,000 active editors (https://www.wikidata.org/wiki/Wikidata:Statistics) and growing, it is able to cover a much wider number of concepts than any user individually.
It is fast, because one does not need to install any software or ask for permissions to update it: any user can simply do it via a web interface.
That speed makes it easier for newcomers to join and contribute, in contrast to OBO Foundry ontologies, which require extensive training on semantics and knowledge of Git/GitHub for contributions.
Finally, the information on Wikidata is available via an user interface, via a SPARQL query service and as large, full-size database dumps, providing full extent reusability.
The Wikidata model has been so sucessfull that Google decided to migrate its own knowledge base, Freebase, fully into Wikidata.{[}\protect\hyperlink{ref-xLpRePoh}{80}{]}

@@

Wikidata uses the same framework (RDF) that powers ontologies, and its model represents statements about the world in triples containing a subjects, a property and an object. {[}\protect\hyperlink{ref-Ea6hVYrD}{81}{]}
Its data model is serialized both in JSON and RDF.
The data model contains 17 different datatypes, including, for example ``Item'', any entry on Wikidata that refers to ``o a real-world object, concept, or event that is given an identifier in Wikidata'' and ``String'', a ``sequence of freely chosen characters interpreted as text''. {[}\protect\hyperlink{ref-17Yt10yrY}{82}{]}.
Knowledge is stored on Wikidata upon basic triples containing a subject (of type ``Item''), a property and a value (which can be of any of the 17 types).
As of November 2021, Wikidata contains more than 90 million data items {[}\protect\hyperlink{ref-SGzjBOG0}{83}{]} and more than 9000 properties that link them to values.
As values often are other items, the database aquires a network format with labeled edges.

As can be seen in the example in \ref{fig:douglas_adams}, each the items in the database contain an item identifiers (Q followed by numbers).
They also contain a label, a description and a list of aliases, which can be recorded in any of the more than 200 hundred languages, thereby making it a multilingual project. {[}\protect\hyperlink{ref-S1Lksw7i}{84}{]}
Each item is decorated with statements, comprising of property-value pairs.
These pairs can be further specified via qualifiers and references, which treats the full triple as the subject, adding metadata to it (a process called reification {[}\protect\hyperlink{ref-ZhP1mKyl}{85}/\#reification{]}).
Qualifiers provide ways to extend the information on the triple, while references provide provenance, enabling users to judge the validity of the claims in the database.

@@

All the information is available on a user interface, but its data is also available programatically in diverse formats, including as full JSON and RDF dumps, the MediaWiki API and a SPARQL endpoint. {[}\protect\hyperlink{ref-4wSLyAPr}{86}{]}
A number of wrappers of such services are available in languages such as R {[}\protect\hyperlink{ref-1HghnypzG}{87}{]} and python {[}\protect\hyperlink{ref-VKWfrtR1}{88}/{]}.
A scheme of the data can be seen in \ref{fig:datamodel}, where each item is connected to a statement node via a property in the ``p:'' namespace, from which references and qualifiers are acessible.
To facilitate basic usage, the namespace ``wdt:'' connects items to values directly, simplifying, for example, the writing of SPARQL queries.

@@

Information on Wikidata is released under a CC0 license, which enables full reuse of the data. {[}\protect\hyperlink{ref-tQIJPG4k}{89}{]}
One of the major points of access and reuse of the information is the Wikidata Query Service {[}\protect\hyperlink{ref-2wDsXBwd}{90}/{]}, a core resource of the community which enables live querying in the SPARQL language. {[}\protect\hyperlink{ref-t66jYhqe}{91}{]}
A number of services make use of embedded queries from the Wikidata Query Service {[}\protect\hyperlink{ref-2wDsXBwd}{90}/{]} to create interactive, live dashboards, for example Scholia {[}\protect\hyperlink{ref-17P4zruFZ}{92}/{]} aht the SARS-CoV-2 Query Book {[}\protect\hyperlink{ref-guMp3q1V}{93}/{]}

Wikidata is not only acessible in different ways, but also writable in many ways.
It provides a user-friendly, point-and-click interface for modifying the database, providing a low entry barriers for newcomers.
It is also possible to semi-automatically reconcile spreadsheets to Wikidata items and use batch tools such as Open Refine {[}\protect\hyperlink{ref-17rRjL1Xj}{94}{]} and Quickstatements {[}\protect\hyperlink{ref-K6ggpwRg}{95}{]}, which enable batches on the magnitude of thousands of edits.
For larger amounts of edits, it is possible to ask for bot permissions {[}\protect\hyperlink{ref-sSH04fxa}{96}{]} and deploy systems that integrate big data sources.
Bot edits are made via the WikiMedia API and are predominantly written via Python wrappers, such as Pywikibot {[}\protect\hyperlink{ref-5K5MfnVl}{97}{]} and the Wikidata Integrator. {[}\protect\hyperlink{ref-qDI8I4IJ}{98}{]}

@@

Due to its privileged position inside the linked data ecosystem and its ease of write and query, Wikidata has been growing as a hub for interoperable data for the life sciences community. {[}\protect\hyperlink{ref-3GqlN9Dk}{65}{]} {[}\protect\hyperlink{ref-Ym2HoRLl}{99}{]}
Even though Wikidata was created in 2013, the demand for a community-cured life sciences knowledge graph is clear at least since 2008 {[}\protect\hyperlink{ref-179mXrGIk}{100}{]} {[}\protect\hyperlink{ref-TItj7EYf}{101}{]}
The Wikidata-like project proposed at the time was eventually discontinued, an example of the challenge of maintaining independent biomedical databases. {[}\protect\hyperlink{ref-P2QdZvGc}{102}{]}
As Wikidata has a very large community, has stable funding and is at the core of modern technologies, like the Google Knowledge Graph {[}\protect\hyperlink{ref-xLpRePoh}{80}{]} and Amazon's Alexa, {[}\protect\hyperlink{ref-bO7BekzO}{103}/{]} it is virtually guaranteed that data in Wikidata will remain acessible for a long time, regardless of local funding schemes.

The Gene Wiki project {[}\protect\hyperlink{ref-18UXATsxP}{104}{]} was likely the first large scale biomedical project to rely directly on the Wikipedia infrastructure for community curation.
It provided a direction connection between the generalist community of Wikipedia and domain experts.
The interplay of both communities is a topic of discussion and the opportunities and challenges were already discussed in NAR in 2012. {[}\protect\hyperlink{ref-13LWXQF0E}{105}{]}\\
Notably, Wikidata appeared chronologically after those efforts.\\
Notwithstandingly, the Gene Wiki research group has embraced the Wikidata environment for community biocuration and data interoperability {[}\protect\hyperlink{ref-2ZhxC0dg}{106}{]}{[}\protect\hyperlink{ref-DaJdrXLk}{107}{]} {[}\protect\hyperlink{ref-3GqlN9Dk}{65}{]} {[}\protect\hyperlink{ref-aQfGxPRH}{108}{]}.
The information on Wikidata is still integrated to Wikipedias across multiple languages, often as source of information in Wikipedia's infoboxes.

Other projects outside the Gene Wiki innitiative also started using Wikidata as a platform for knowledge integration.
A list of several projects that use Wikidata as part of their service to their community is given on table 1.
There is movement exploring how Wikidata can be employed to the advance of Computational Biology, and how it can be integrated to current publication status quo. {[}\protect\hyperlink{ref-1GijvCQtv}{109}{]}
In that direction, Wikidata is being developed as a platform for scholarly linked open data, particularly via the Scholia platform {[}\protect\hyperlink{ref-hxzL9pmm}{110}{]} {[}\protect\hyperlink{ref-PdG2G4Bp}{111}{]},(\url{https://scholia.toolforge.org/}) which provides profiles of pre-templated SPARQL queries for entities like particular authors and articles (e.g.~Scholia profile on Prof.~Helder Nakaya: https://scholia.toolforge.org/author/Q42614737).

@@

During the COVID-19 pandemic, Wikidata has spawned as a hotspot for modelling information about the virus and the pandemic in real time. {[}\protect\hyperlink{ref-zH3KDqhp}{112}{]} {[}\protect\hyperlink{ref-wikidata:99196713}{\textbf{wikidata:99196713?}}{]}
The general scope of the databasae allowed representation in a shared system of molecular, epidemiologic and socio-economic aspects of the pandemic. {[}\protect\hyperlink{ref-zH3KDqhp}{112}{]}{[}\protect\hyperlink{ref-mPoPwN77}{113}{]}
Information curated in Wikidata was immediately available, feeding live dashboards and other applications based on SPARQL queries .{[}\protect\hyperlink{ref-3YCzr3jU}{114}{]} {[}\protect\hyperlink{ref-SkOeuFRd}{115}{]} {[}\protect\hyperlink{ref-AOVCR8mW}{116}{]}
Additionally, as the information presented on Wikidata is multilingual and collaboratively edited, it presented itself as a resource for constructing structured vocabularies in non-english languages. {[}\protect\hyperlink{ref-19wiU1LCA}{117}{]}

>>>>

This manuscript
(\href{https://lubianat.github.io/quali_phd/v/e6cf449283aa647015cb56637d7c2de7d9f53a02/}{permalink})
was automatically generated
from \href{https://github.com/lubianat/quali_phd/tree/e6cf449283aa647015cb56637d7c2de7d9f53a02}{lubianat/quali\_phd@e6cf449}
on December 7, 2021.

@@

The Human Cell Atlas (HCA) is an international effort aiming at characterizing every cell type of the human body.
By the virtue of tecniques such as single-cell RNA sequencing, mass cytometry, and multiplexed in situ hybridization, HCA members are producing cell-level data from virtually all human tissues.
This wealth of data can have a significant impact on biomedical research, but only if its content is genuinely interoperable.
While ontologies and semantic technologies have emerged as key players in the data interoperability ecosystem, there are still gaps to cover between the technical possibilities and the practical applications in biomedical research.
In addition to ontologies, like the Cell Ontology and the Gene Ontology, large-scale knowledge graphs are growing as a tool for knowledge makanagement.
Among those, Wikidata, a sister project of Wikipedia for structured data, is surfacing as a hub in the semantic web for multiple types of information.
The formatting and deployment of information from the Human Cell Atlas to Wikidata can increase information availability and impact, conecting the scientific products with the larger knowledge ecosystem.
This PhD project aims at studying Wikidata as a platfomr for representing cell types, adressing theoretical and pratical concerns.\\
We are reviewing the literature on cell types, refining and formalizing concepts for cell type delimitation.
At the same time, we are enriching Wikidata with new classes curated from the literature, and with large scale integrations of biomedical databases (e.g.~PanglaoDB) into the Wikidata infrastructure.
To aid that effort, we are developing Wikidata Bib, a framework for literature management and organized note-taking and recording system for reading the academic literature with high efficiency.
Finally, we plan to improve the interplay of Wikidata, the Cell Ontology and software used for single-cell RNA-seq data, inserting Wikidata \emph{de facto} as a tool for the Human Cell Atlas community.

\hypertarget{preface}{%
\section{Preface}\label{preface}}

Here we present an overview of the different chapters that compose this document, presented as the text for a qualifying exam.\\
This work is concerned with the conceptual modelling of knowledge about cell types.
The introduction contains an overview of the Human Cell Atlas project, and the current state of classification of cells into types.
Then, it proceeds to introduce ontologies and knowledge graphs as tools for connecting what we know about cells.

The methodology section is an overview of the core methods used throughout the work.
However, as the project contains elements from different scientific traditions, the results chapters might also display particular methods used in the specific branch of the project.

It is worth noticing that the different results shown were not developed chronologically in the order shown.
They were actually developed in parallel, with overlapping periods of activity.
They have been organized into separate chapters, however, as they tackle different perspectives of the subject matter, and are part of different publications.

The discussion on the concept of cell type is presented first, as it is instrumental for the later steps.
It is followed by an account of how PanglaoDB, a database of cell markers, was integrated to Wikidata, based on a notion of species-specific cell type clarified in the preceding chapter.

Then, we present Wikidata Bib, a framework for organized reading of the literature.
The framework, althoug used as a method throughout the PhD project, is presented in the results session.
This emphasis as a result was chosen as the technical and theoretical details of the system are part of the intelectual work put into the project.
The system evolved into a biocuration platform for the collection of cell types from the literature to Wikidata, and the statistics on this curation are also presented on the section.

To end the results, we discuss how our efforts integrate with the Cell Ontology, the currently leading system for organizing cell types.

Finally, an account of other academic aspects of the project are presented, as part of the qualification requirements.
They present an overview of collaborations, participation in events and academic courses taken during the first part of the PhD project.

\hypertarget{background}{%
\section{Background}\label{background}}

@@

@@

The classification of biological concepts is at the core of biology. At least since the Aristotelian endeavours to group classes of animals, a good part of the scientific work is to capture concepts into knowledge systems {[}\protect\hyperlink{ref-DLTMbf54}{60}{]}.
Linnaeus' binomial system for naming species and Mendeleiev's periodic table are likely the two most famous classification systems, but are part of a much larger ecosystem of structuring scientific knowledge.

On the 20th century, the development of the analytical philosophy of Russel and Wittgenstein and their search for formalizations {[}\protect\hyperlink{ref-cZEmYsqx}{61}{]} gradually layed the foundations for the the logic of scientific descriptions. Karl Popper and his ``The Logic of Scientific Discovery''{[}\protect\hyperlink{ref-to2onCDy}{62}{]} was heavily influenced by analytical philosophy, and the field is at the foundation of the ``falseability'' system of Popper. Less known among life scientists, Tarski's inquiries on what can be considered to be ``true'' {[}\protect\hyperlink{ref-RghcuiSS}{63}{]} were also

@@

An ontology, as used here, is a formal computational representation of reality, which tries to represent each concept (and their relations) as precisely as possible. {[}\protect\hyperlink{ref-DLTMbf54}{60}{]}\\
Constructing an ontology is a process of selecting and defining terms of interest, selecting and defining relationships of interest and making statements about reality using terms and relationships.
The Gene Ontology is probably the most well known biomedical ontology; it describes (among other things) different classes of biological process, related to each\_other by ``is\_a'' and ``part\_of relations. {[}\protect\hyperlink{ref-18wkLcGxT}{64}{]} {[}\protect\hyperlink{ref-2KmDs8gy}{65}{]}.

The Gene Ontology is part of a much larger effort to formalize concepts across biology: the Open Biomedical and Biological Ontologies (OBO) Foundry. {[}\protect\hyperlink{ref-j3tJyG7J}{66}{]}
Created in 2007, the OBO Foundry is a hub of biomedical ontologies that sets guidelines for the design and construction of high-quality ontologies.
The initial OBO Foundry united several independent ontologies, like the Cell Ontology (CL), the Disease Ontology (DO) and the Protein Ontology (PRO) under a common framework towards interoperability.
At the same time, the creation of the Relation Ontology (RO) provided a go-to point for relations in biology that could them be reused by different ontologies.

@@

The Google Knowledge Graph introduced the Semantic Web \emph{de facto} in the daily life of users of Google. {[}\protect\hyperlink{ref-16mCkZC1r}{67}/{]}. Its underlying structure is similar to the triples in an ontology, but it is less concerned with being logically coherent, and does have strict semantics of a representation.
In that way, Google Knowledge Graphs can feed on a variety of sources and not crash if there is some data modelling that, rigorously, could be inconsistent.
Even though there is not a rigorous boundary between ontologies and knowledge graphs, one reasonable interpretation is that a knowledge graph may not be perfectly coherent, as long as it still can provide enough knowledge and reasoning for the approach of interest.
While the lack of formal semantics limits reasoning and inference, the knowledge graphs are arguably easier to use, edit and and understand, and so provide an user friendly alternative for computable information with a lower entry barrier.

While the Google Knowledge Graph is widely used as a source of knowledge, it does not allow independent users to contribute with information.
On the other hand, Wikidata, the collaborative knowledge graph of the Wikimedia foundation, allows users to contribute with classes and statements, in the same spirit of Wikipedia and share its ``epistemic virtues, like power, speed and availability. {[}\protect\hyperlink{ref-qMozvNth}{68}{]}
Its power is derived of its large community of contributors, closedly linked to the hugely successful Wikipedia.
With a community of more than 20,000 active editors (https://www.wikidata.org/wiki/Wikidata:Statistics) and growing, it is able to cover a much wider number of concepts than any user individually.
It is fast, because one does not need to install any software or ask for permissions to update it: any user can simply do it via a web interface.
That speed makes it easier for newcomers to join and contribute, in contrast to OBO Foundry ontologies, which require extensive training on semantics and knowledge of Git/GitHub for contributions.
Finally, the information on Wikidata is available via an user interface, via a SPARQL query service and as large, full-size database dumps, providing full extent reusability.
The Wikidata model has been so sucessfull that Google decided to migrate its own knowledge base, Freebase, fully into Wikidata.{[}\protect\hyperlink{ref-xLpRePoh}{69}{]}

@@

Wikidata uses the same framework (RDF) that powers ontologies, and its model represents statements about the world in triples containing a subjects, a property and an object. {[}\protect\hyperlink{ref-Ea6hVYrD}{70}{]}
Its data model is serialized both in JSON and RDF.
The data model contains 17 different datatypes, including, for example ``Item'', any entry on Wikidata that refers to ``o a real-world object, concept, or event that is given an identifier in Wikidata'' and ``String'', a ``sequence of freely chosen characters interpreted as text''. {[}\protect\hyperlink{ref-17Yt10yrY}{71}{]}.
Knowledge is stored on Wikidata upon basic triples containing a subject (of type ``Item''), a property and a value (which can be of any of the 17 types).
As of November 2021, Wikidata contains more than 90 million data items {[}\protect\hyperlink{ref-SGzjBOG0}{72}{]} and more than 9000 properties that link them to values.
As values often are other items, the database aquires a network format with labeled edges.

As can be seen in the example in \ref{fig:douglas_adams}, each the items in the database contain an item identifiers (Q followed by numbers).
They also contain a label, a description and a list of aliases, which can be recorded in any of the more than 200 hundred languages, thereby making it a multilingual project. {[}\protect\hyperlink{ref-S1Lksw7i}{73}{]}
Each item is decorated with statements, comprising of property-value pairs.
These pairs can be further specified via qualifiers and references, which treats the full triple as the subject, adding metadata to it (a process called reification {[}\protect\hyperlink{ref-ZhP1mKyl}{74}/\#reification{]}).
Qualifiers provide ways to extend the information on the triple, while references provide provenance, enabling users to judge the validity of the claims in the database.

@@

All the information is available on a user interface, but its data is also available programatically in diverse formats, including as full JSON and RDF dumps, the MediaWiki API and a SPARQL endpoint. {[}\protect\hyperlink{ref-4wSLyAPr}{75}{]}
A number of wrappers of such services are available in languages such as R {[}\protect\hyperlink{ref-1HghnypzG}{76}{]} and python {[}\protect\hyperlink{ref-VKWfrtR1}{77}/{]}.
A scheme of the data can be seen in \ref{fig:datamodel}, where each item is connected to a statement node via a property in the ``p:'' namespace, from which references and qualifiers are acessible.
To facilitate basic usage, the namespace ``wdt:'' connects items to values directly, simplifying, for example, the writing of SPARQL queries.

@@

Information on Wikidata is released under a CC0 license, which enables full reuse of the data. {[}\protect\hyperlink{ref-tQIJPG4k}{78}{]}
One of the major points of access and reuse of the information is the Wikidata Query Service {[}\protect\hyperlink{ref-2wDsXBwd}{79}/{]}, a core resource of the community which enables live querying in the SPARQL language. {[}\protect\hyperlink{ref-t66jYhqe}{80}{]}
A number of services make use of embedded queries from the Wikidata Query Service {[}\protect\hyperlink{ref-2wDsXBwd}{79}/{]} to create interactive, live dashboards, for example Scholia {[}\protect\hyperlink{ref-17P4zruFZ}{81}/{]} aht the SARS-CoV-2 Query Book {[}\protect\hyperlink{ref-guMp3q1V}{82}/{]}

Wikidata is not only acessible in different ways, but also writable in many ways.
It provides a user-friendly, point-and-click interface for modifying the database, providing a low entry barriers for newcomers.
It is also possible to semi-automatically reconcile spreadsheets to Wikidata items and use batch tools such as Open Refine {[}\protect\hyperlink{ref-17rRjL1Xj}{83}{]} and Quickstatements {[}\protect\hyperlink{ref-K6ggpwRg}{84}{]}, which enable batches on the magnitude of thousands of edits.
For larger amounts of edits, it is possible to ask for bot permissions {[}\protect\hyperlink{ref-sSH04fxa}{85}{]} and deploy systems that integrate big data sources.
Bot edits are made via the WikiMedia API and are predominantly written via Python wrappers, such as Pywikibot {[}\protect\hyperlink{ref-5K5MfnVl}{86}{]} and the Wikidata Integrator. {[}\protect\hyperlink{ref-qDI8I4IJ}{87}{]}

@@

Due to its privileged position inside the linked data ecosystem and its ease of write and query, Wikidata has been growing as a hub for interoperable data for the life sciences community. {[}\protect\hyperlink{ref-3GqlN9Dk}{88}{]} {[}\protect\hyperlink{ref-Ym2HoRLl}{89}{]}
Even though Wikidata was created in 2013, the demand for a community-cured life sciences knowledge graph is clear at least since 2008 {[}\protect\hyperlink{ref-179mXrGIk}{90}{]} {[}\protect\hyperlink{ref-TItj7EYf}{91}{]}
The Wikidata-like project proposed at the time was eventually discontinued, an example of the challenge of maintaining independent biomedical databases. {[}\protect\hyperlink{ref-P2QdZvGc}{92}{]}
As Wikidata has a very large community, has stable funding and is at the core of modern technologies, like the Google Knowledge Graph {[}\protect\hyperlink{ref-xLpRePoh}{69}{]} and Amazon's Alexa, {[}\protect\hyperlink{ref-bO7BekzO}{93}/{]} it is virtually guaranteed that data in Wikidata will remain acessible for a long time, regardless of local funding schemes.

The Gene Wiki project {[}\protect\hyperlink{ref-18UXATsxP}{94}{]} was likely the first large scale biomedical project to rely directly on the Wikipedia infrastructure for community curation.
It provided a direction connection between the generalist community of Wikipedia and domain experts.
The interplay of both communities is a topic of discussion and the opportunities and challenges were already discussed in NAR in 2012. {[}\protect\hyperlink{ref-13LWXQF0E}{95}{]}\\
Notably, Wikidata appeared chronologically after those efforts.\\
Notwithstandingly, the Gene Wiki research group has embraced the Wikidata environment for community biocuration and data interoperability {[}\protect\hyperlink{ref-2ZhxC0dg}{96}{]}{[}\protect\hyperlink{ref-DaJdrXLk}{97}{]} {[}\protect\hyperlink{ref-3GqlN9Dk}{88}{]} {[}\protect\hyperlink{ref-aQfGxPRH}{98}{]}.
The information on Wikidata is still integrated to Wikipedias across multiple languages, often as source of information in Wikipedia's infoboxes.

Other projects outside the Gene Wiki innitiative also started using Wikidata as a platform for knowledge integration.
A list of several projects that use Wikidata as part of their service to their community is given on table 1.
There is movement exploring how Wikidata can be employed to the advance of Computational Biology, and how it can be integrated to current publication status quo. {[}\protect\hyperlink{ref-1GijvCQtv}{99}{]}
In that direction, Wikidata is being developed as a platform for scholarly linked open data, particularly via the Scholia platform {[}\protect\hyperlink{ref-hxzL9pmm}{100}{]} {[}\protect\hyperlink{ref-PdG2G4Bp}{101}{]},(\url{https://scholia.toolforge.org/}) which provides profiles of pre-templated SPARQL queries for entities like particular authors and articles (e.g.~Scholia profile on Prof.~Helder Nakaya: https://scholia.toolforge.org/author/Q42614737).

@@

During the COVID-19 pandemic, Wikidata has spawned as a hotspot for modelling information about the virus and the pandemic in real time. {[}\protect\hyperlink{ref-zH3KDqhp}{102}{]} {[}\protect\hyperlink{ref-wikidata:99196713}{\textbf{wikidata:99196713?}}{]}
The general scope of the databasae allowed representation in a shared system of molecular, epidemiologic and socio-economic aspects of the pandemic. {[}\protect\hyperlink{ref-zH3KDqhp}{102}{]}{[}\protect\hyperlink{ref-mPoPwN77}{103}{]}
Information curated in Wikidata was immediately available, feeding live dashboards and other applications based on SPARQL queries .{[}\protect\hyperlink{ref-3YCzr3jU}{104}{]} {[}\protect\hyperlink{ref-SkOeuFRd}{105}{]} {[}\protect\hyperlink{ref-AOVCR8mW}{106}{]}
Additionally, as the information presented on Wikidata is multilingual and collaboratively edited, it presented itself as a resource for constructing structured vocabularies in non-english languages. {[}\protect\hyperlink{ref-19wiU1LCA}{107}{]}
